Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2018-09-06 paper/tidy-missing-data-paper.Rmd
* Added JCGS CSL * updated gitignore to include preamble * updated preamble with pieces from Rob's Monash template

---
title: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
author:
- familyname: Tierney
  othernames: Nicholas
  address: Monash University
  email: nicholas.tierney@gmail.com
  correspondingauthor: true
- familyname: Cook
  othernames: Dianne
  address: Monash University
abstract: "Despite the large body of research on missing value distributions and imputation, there is comparatively little literature on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with a goal to integrating missing value handling as an integral part of data analysis workflows. New data structures are defined along with new functions (verbs) to perform common operations. Together these provide a cohesive framework for handling, exploring, and imputing missing values. These methods have been made available in the R package `naniar`."
keywords: "workflow, statistical computing, data science, data visualization, tidyverse, data pipeline"
wpnumber: 1/2018
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: tidy-missing-data-paper.bib
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
---

@@

The approach of @Unwin1996 focussed on multivariate categorical data, where missingness is explicitly added as an additional category. MANET also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot. The approach of @Swayne1990 in the software XGobi, further developed in ggobi [@Cook2007], focussed more on multivariate quantitative data. Missingness is incorporated into plots in ggobi by setting them to be 10% below the minimum value.

@@

```{r rpart-plot, fig.height = 2, fig.width = 4, out.width = "100%", fig.align = "center", fig.cap = "Decision tree output predicting the clusters of missingness. Type of house, the year quarter, and year were important for predicting missingness cluster. The cluster with the most missingness was for quarters 1 and 4, for 2017 and 2018. Type of house, year, and year quarter are important features related to the missingness structure."}

>>>>

---
title: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
author: "Nicholas Tierney, Department of Econometrics and Business Statistics, Monash University, **Corresponding author (nicholas.tierney@gmail.com)** \\newline Dianne Cook, Department of Econometrics and Business Statistics, Monash University"
date: "September 6, 2018"
toc: false
bibliography: tidy-missing-data-paper.bib
output:
  bookdown::pdf_document2:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: yes
    number_sections: yes
csl: jcgs.csl
---

**Abstract**:

Despite the large body of research on missing value distributions and imputation, there is comparatively little literature on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with a goal to integrating missing value handling as an integral part of data analysis workflows. New data structures are defined along with new functions (verbs) to perform common operations. Together these provide a cohesive framework for handling, exploring, and imputing missing values. These methods have been made available in the R package `naniar`.

_Keywords:_ workflow, statistical computing, data science, data visualization, tidyverse, data pipeline

@@

The approach of @Unwin1996 focussed on multivariate categorical data, where missingness is explicitly added as an additional category. MANET also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot. The approach of @Swayne1998 in the software XGobi, further developed in ggobi [@Cook2007], focussed more on multivariate quantitative data. Missingness is incorporated into plots in ggobi by setting them to be 10% below the minimum value.

@@

```{r rpart-plot, fig.height = 4, fig.width = 8, out.width = "100%", fig.align = "center", fig.cap = "Decision tree output predicting the clusters of missingness. Type of house, the year quarter, and year were important for predicting missingness cluster. The cluster with the most missingness was for quarters 1 and 4, for 2017 and 2018. Type of house, year, and year quarter are important features related to the missingness structure."}
